```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are 3-4 packages you will need to install for today's practical: `install.packages(c("xgboost", "eegkit", "forecast", "tseries", "caret"))` apart from that everything else should already be available on your system. 

If you are using a newer Mac you may have to also install [quartz](https://www.xquartz.org/) to have everything work (do this if you see errors about `X11` during install/execution).

I will endeavour to use explicit imports to make it clear where functions are coming from (functions without `library_name::` are part of base R or a function we've defined in this notebook).

```{r libraries, echo=FALSE}
# Using the same library we used earlier in the course for tabular data because we know it works!
library(xgboost)

# EEG manipulation library in R (although very limited compared to signal processing libraries available in other languages, matlab might actually still be a leader in this specific area)
library(eegkit)

# some time series functions (that we only skim the depths of)
library(forecast)
library(tseries)
library(caret)

# just tidyverse libraries that should already be installed
library(dplyr)
library(reshape2)
library(purrr)
library(ggplot2)
```

## EEG Eye Detection Data

One of the most common types of medical sensor data (and one that we talked about during the lecture) are Electroencephalograms (EEGs).  
These measure mesoscale electrical signals (measured in microvolts) within the brain, which are indicative of a region of neuronal activity.
Typically, EEGs involve an array of sensors (aka channels) placed on the scalp with a high degree of covariance between sensors.

As EEG data can be very large and unwieldy, we are going to use a relatively small/simple dataset today from [this paper](http://ehrai.com/su/pdf/aihls2013.pdf).

This dataset is a 117 second continuous EEG measurement collected from a single person with a device called a "Emotiv EEG Neuroheadset".
In combination with the EEG data collection, a camera was used to record whether person being recorded had their eyes open or closed. 
This was eye status was then manually annotated onto the EEG data with `1` indicated the eyes being closed and `0` the eyes being open.
Measures microvoltages are listed in chronological order with the first measured value at the top of the dataframe.

Let's parse the data directly from the `h2o` library's (which we aren't actually using directly) test data S3 bucket:

```{r parse_data}
eeg_url <- "https://h2o-public-test-data.s3.amazonaws.com/smalldata/eeg/eeg_eyestate_splits.csv"
eeg_data <- read.csv(eeg_url)

# add timestamp
Fs <- 117 / nrow(eeg_data)
eeg_data <- transform(eeg_data, ds = seq(0, 116.99999, by = Fs), eyeDetection = as.factor(eyeDetection))
print(table(eeg_data$eyeDetection))

# split dataset into train, validate, test
eeg_train <- subset(eeg_data, split == 'train', select = -split)
print(table(eeg_train$eyeDetection))

eeg_validate <- subset(eeg_data, split == 'valid', select = -split)
eeg_test <- subset(eeg_data, split == 'test', select = -split)
```

**0** Knowing the `eeg_data` contains 117 seconds of data, inspect the `eeg_data` dataframe and the code above to and determine how many samples per second were taken?

1/Fs = 1/0.00781041388518024 = 128.0342

**1** How many EEG electrodes/sensors were used?

There were 14 EEG electrodes/sensors used, as there are 14 variables that represent the electrodes.

### Exploratory Data Analysis

Now that we have the dataset and some basic parameters let's begin with the ever important/relevant exploratory data analysis.

First we should check there is no missing data!
```{r check_na}
sum(is.na(eeg_data))
```

Great, now we can start generating some plots to look at this data within the time-domain.

First we use `reshape2::melt()` to transform the `eeg_data` dataset from a wide format to a long format expected by `ggplot2`.

Specifically, this converts from "wide" where each electrode has its own column, to a "long" format, where each observation has its own row. 
This format is often more convenient for data analysis and visualization, especially when dealing with repeated measurements or time-series data.

We then use `ggplot2` to create a line plot of electrode intensities per sampling time, with the lines coloured by electrode, and the eye status annotated using dark grey blocks.

```{r plot_data}
melt <- reshape2::melt(eeg_data %>% dplyr::select(-split), id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")


ggplot2::ggplot(melt, ggplot2::aes(x=ds, y=microvolts, color=Electrode)) + 
  ggplot2::geom_line() + 
  ggplot2::ylim(3500,5000) + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(melt, eyeDetection==1), alpha=0.005)
```

**2** Do you see any obvious patterns between eyes being open (dark grey blocks in the plot) and the EEG intensities?

When the eyes are open (dark grey blocks) there are usually upward spikes on the EEG intensity at the beginning of each dark grey blocks. Throughout the lighter grey blocks the EEG intensities seem to be more uniform with smaller spikes, when the eyes are closed. 

**3** Similarly, based on the distribution of eye open/close state over time do you anticipate any temporal correlation between these states?

**do this later**

Let's see if we can directly look at the distribution of EEG intensities and see how they related to eye status.


As there are a few extreme outliers in voltage we will use the `dplyr::filter` function to remove values outwith of 3750 to 50003. The function uses the `%in%` operator to check if each value of microvolts is within that range. The function also uses the `dplyr::mutate()` to change the type of the variable eyeDetection from numeric to a factor (R's categorical variable type).

```{r compare_distrib}
melt_train <- reshape2::melt(eeg_train, id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")

# filter huge outliers in voltage
filt_melt_train <- dplyr::filter(melt_train, microvolts %in% (3750:5000)) %>% dplyr::mutate(eyeDetection=as.factor(eyeDetection))

ggplot2::ggplot(filt_melt_train, ggplot2::aes(y=Electrode, x=microvolts, fill=eyeDetection)) + ggplot2::geom_boxplot()
```

Plots are great but sometimes so it is also useful to directly look at the summary statistics and how they related to eye status.
We will do this by grouping the data based on eye status and electrode before calculating the statistics using the convenient `dplyr::summarise` function.

```{r compare_summary_stats}
filt_melt_train %>% dplyr::group_by(eyeDetection, Electrode) %>% 
    dplyr::summarise(mean = mean(microvolts), median=median(microvolts), sd=sd(microvolts)) %>% 
    dplyr::arrange(Electrode)

```

**4** Based on these analyses are any electrodes consistently more intense or varied when eyes are open?

Based on these analyses there are two electrodes (T8 and F8) that are slightly more intense when the eyes are closed, since the means  and medians are larger for when EyeDetection=1 (eyes closed). 

Based on these analyses there are two electrodes (P7 and O1) that are more varied when the eyes are closed as the standard deviations for these are larger for EyeDetection=1 (eyes closed). 

#### Time-Related Trends

As it looks like there may be a temporal pattern in the data we should investigate how it changes over time.  

First we will do a statistical test for stationarity:

```{r convert_to_tseries}
apply(eeg_train, 2, tseries::adf.test)
```

**5** What is stationarity?

Stationarity is used in time series data and shows that means, medians, standard deviations, and autocorrelations do not change over time. 

**6** Why are we interested in stationarity? What do the results of these tests tell us? (ignoring the lack of multiple comparison correction...)

We are interested in stationarity because we want to see if the means and variances of the EEG intensities change over time and comparing when the eyes are open versus when the eyes are closed. 

All of the p-values are are significant except for the last one (0.4045), therefore the alternative hypotheses were accepted and the data is stationary. 

Then we may want to visually explore patterns of autocorrelation (previous values predicting future ones) and cross-correlation (correlation across channels over time) using `forecast::ggAcf` function.

The ACF plot displays the cross-correlation between each pair of electrode channels and the auto-correlation within the same electrode (the plots along the diagonal.)

Positive autocorrelation indicates that the increase in voltage observed in a given time-interval leads to a proportionate increase in the lagged time interval as well.
Negative autocorrelation indicates the opposite!


```{r correlation}
forecast::ggAcf(eeg_train %>% dplyr::select(-ds))
```

**7** Do any fields show signs of strong autocorrelation (diagonal plots)? Do any pairs of fields show signs of cross-correlation? Provide examples.

The fields that show signs of strong autocorrelation are: FC5, F7, FC6, O1

The pairs of fields that show signs of cross-correlation are: FC5 and F7

#### Frequency-Space 

We can also explore the data in frequency space by using a Fast Fourier Transform.  
After the FFT we can summarise the distributions of frequencies by their density across the power spectrum.
This will let us see if there any obvious patterns related to eye status in the overall frequency distributions.

```{r fft_open}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 0) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Open")
```

```{r fft_closed}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 1) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Closed")
```

**8** Do you see any differences between the power spectral densities for the two eye states? If so, describe them.

**finish later**

Eyes open: less blue, more red at the top (channel 14), orange in the middle (channel 6)

Eyes closed: more blue, more orange/red at the bottom (channel 1 and 9), dark yellow at top (channel 14)

#### Independent Component Analysis

We may also wish to explore whether there are multiple sources of neuronal activity being picked up by the sensors.  
This can be achieved using a process known as independent component analysis (ICA) which decorrelates the channels and identifies the primary sources of signal within the decorrelated matrix.

```{r ica, warning=FALSE}
ica <- eegkit::eegica(eeg_train %>% dplyr::select(-eyeDetection, -ds), nc=3, method='fast', type='time')
mix <- dplyr::as_tibble(ica$M)
mix$eyeDetection <- eeg_train$eyeDetection
mix$ds <- eeg_train$ds

mix_melt <- reshape2::melt(mix, id.vars=c("eyeDetection", "ds"), variable.name = "Independent Component", value.name = "M")


ggplot2::ggplot(mix_melt, ggplot2::aes(x=ds, y=M, color=`Independent Component`)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(mix_melt, eyeDetection==1), alpha=0.005) +
  ggplot2::scale_y_log10()
```



**9** Does this suggest eye opening relates to an independent component of activity across the electrodes?

**FINISH**

Dark grey: eyes open - when eyes open there is a big spike downwards - therefore this does suggest that eye opening relates to an independent component of activity across the electrodes.

### Eye Opening Prediction

Now that we've explored the data let's use a simple model to see how well we can predict eye status from the EEGs:

```{r xgboost}
# Convert the training and validation datasets to matrices
eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) -1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) -1

# Build the xgboost model
model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

print(model)
```


**10** Using the `caret` library (or any other library/model type you want such as a neural network) fit another model to predict eye opening.

```{r}
# training and test set

data.prep <- eeg_data 
```

```{r model2}
# Build the caret model
fit.control <- trainControl(method="repeatedcv", number=6,repeats=6)
crt.tree <- train(eyeDetection~., data=eeg_train, method="rpart", trControl=fit.control)
crt.tree

print(crt.tree)
```


**11** Using the best performing of the two models (on the validation dataset) calculate and report the test performance (filling in the code below):


```{r test}
confusionMatrix()
```

**12** Describe 2 possible alternative modelling approaches for prediction of eye opening from EEGs we discussed in class but haven't explored in this notebook.

Frequency/Spectral domain: fourier transform - library gsignal

Path representation: State-space models - independent component analysis - library statespacer 

**13** Find 2 R libraries you could use to implement these approaches.

gsignal and statespacer


### Optional 

**14** (Optional) As this is the last practical of the course - let me know how you would change future offerings of this course.  What worked and didn't work for you (e.g., in terms of the practicals, tutorials, and lectures)? What would you add or remove from the course? What was the main thing you will take away from this course? This will not impact your marks!
